Supplementary Materials

### 2023-01-18 manuscript/manuscript.tex
ms updates

>>>>

\subsection*{Evolutionary Models}

CDS+phastcons+UTR model

 - UTR as neutral control

 - CDS takes priority over phastcons, phastcons is fallback. Otherwise, the
   effect of CDS would be biased by the fraction removed and classified as
   phastcons. This is the right comparison for the substitution model.

\subsection*{Predicted Substitution Rates}

- mention neg. auto correlation?

- Gene level analysis

To evaluate the predicted substitution rate of our CDS+phastcons+UTR model,

Additionally, our GBGS model makes predictions about the rate of substitutions
in different segments under purifying selection. There are two stages of the
prediction process. First, when the initial B' maps are calculated, the
predicted substitution rate is calculated as a byproduct of

\section*{Methods}

\subsection*{Solving the B' Equations for each Segment}

In our implementation of the theory of \textcite{Santiago2016-mu}, we calculate
the local effective population size $N_e(x) = b(x) N$ within a segment $x$ by
solving a system of two non-linear equations,

\begin{align}
  \label{eq:}
  {N}_{e}(x) &= \exp \left( -\frac{V(x)}{2} Q^2(x) \right) & \text{\emph{effective population equation}} \\
  {R}(x) &= \frac{4NU s}{\exp(4 N s)-1}  & \text{\emph{ratchet equation}} 
\end{align}

where,

\begin{align}
  V(x) &= U s - 2 {R}(x) s & \text{\emph{variation equation}} \\
  Q^2(x) &= \frac{2}{(1-Z)(2-(2-M)Z)} & \text{\emph{linkage persistence factor}} \\
  Z &= 1 - \frac{Us}{U-{R(x)}} & \text{\emph{}} \text{\emph{variance decay rate}}
\end{align}

and $U = \mu L$ and $M = r_\text{BP} L$ are the total mutation and
recombination rates for a segment $L$ basepairs long. Each mutation has a
selection coefficient of $s$, for which we assume additive effects. Then, the
local reduction factor is $b(x) = \nicefrac{N_e(x)}{N}$, and the per-basepair
per-generation substitution rate is $\lambda_D = \nicefrac{R(x)}{L}$.

In our software \texttt{bgspy}, these two equations are solved for a grid of
mutation rates and selection coefficients for each specified segment from a
feature type (e.g. CDS, UTRs, etc.). This results in estimates of $N_e(x)$ and
$R(x)$ per segment. In the first-round B' calculation, we use the
drift-effective population size, $N$.

\subsection*{The B' Maps}

The reduction at a focal site $x$ can be calculated as the product of all
segments' contributing reduction factors,

\begin{align}
  B(v) = \prod_x b(x,v)
\end{align}

where $b(x, v)$ is a modified version of $b(x)$ that accounts for the
recombination between the segment $x$ and the neutral site $v$ (see XXX).

In practice, we calculate B' using the variables determined by solving the B'
equations for each mutation and selection coefficient in the grids. Since only
the analog of $Q^2(x)$, $Q^2(x, v)$ changes, the pre-computed substitution rate
(and thus local variance) can be re-used for each $v$. We calculate $B(v)$
across a grid of focal points along the chromosome, spaced 10kb apart.

\subsection*{Maximum Likelihood Estimation}

\subsection*{Local Rescaling and Substitution Rate Prediction}

### 2023-01-18 manuscript/supplementary.tex
ms intro
ms updates

\begin{align}
  \label{suppeqn:b}
  B(x, \Psi) = \prod_g \int_0^1 b(\mu(s, k(g)), s, L_g, r_g, d(x, g)) ds
\end{align}

where, $b(\mu(s, k(g)), s, L_g, r_g, d(x,g))$ is the reduction due to segment
$g$. Here, $\mu(s, k(g))$ is the per-basepair per-genome rate that mutations
with selection coefficient $s$ enter segments with feature class $k(g)$, $L_g$
and $r_g$ are the length in basepairs and recombination rate per basepair of
segment $g$, and the recombination distance between focal site $x$ and segment
$g$ is $d(x, g)$. The functional form of $b$ varies depending on whether the
classic BGS model is used, or our form of Santiago and Caballero's
(\citeyear{Santiago2016-mu}) equations (see Section XXX).

@@

\begin{align}
  {N}_{e,\infty} &= \exp \left( -\frac{V}{2} Q_\infty^2 \right) \\
  {R} &= \frac{4 U s N_{e,\infty}}{\exp(4 N_{e,\infty} s)-1}\\
\end{align}

@@

and $U = 2\mu L$ and $M = r_\text{BP} L$ as above. Solving this system of two
equations for each segment also gives the total substitution or ``ratchet" rate
per generation per segment ($\widetilde{R}$). The per-basepair substitution
rate of the segment is $\lambda = \nicefrac{\widetilde{R}}{L}$

@@

In practice, to fit such a model, we must discretize both $\mu(s, k(g))$ and
$s$. Our models use a fixed $q$-element grid of selection coefficients, and
mutation rates are on a fixed $z$-element grid, which is later interpolated
over during the maximum likelihood optimization. The number of annotation
features $K$ is variable and set by the annotation data specified. There are
two ways to parameterize the mutation rates and DFE for all annotation classes
in background selection models. First, for the \emph{free-mutation}
parameterization, each annotation feature class has a free mutation rate
parameter for each of these selection coefficients (as in equation
\eqref{suppeqn:b}).

In the free-mutation parameterization, there are $1 + K \times q$ for $\pi_0$
and the free mutation rate parameters, where $K$ is the number of annotation
features. The MLE optimization for this approach is unconstrained, though
$\pi_0$ is bounded to be within $10^{-5} \le \pi_0 \le 10^{-1}$ and mutation
rates are bounded within $10^{-11} \le \mu \le 10^{-7}$. The BGS model requires
a mutation rate for each selection coefficient in the grid, for each annotation
class (here, as an example, CDS, UTRs, and phastcons), which are stored in the
$q \times K$ matrix $\mathbf{M}_\text{F}$,

@@

which has $2 + K \times (q-1)$ free parameters, since each column must sum to
one. This imposes the constraint that $\sum_i w_{i,k} = 1$, requiring
constrained MLE optimization.

@@

\begin{align}
  R = \int 
\end{align}

>>>>

\begin{align}
  \label{suppeqn:b}
  B(x, \Psi) = \prod_g^S \int_0^1 b(\mu(s, k(g)), s, L_g, r_g, d(x, g)) ds
\end{align}

where, $b(\mu(s, k(g)), s, L_g, r_g, d(x,g))$ is the reduction due to segment
$g$, and there are $S$ total segments. Here, $\mu(s, k(g))$ is the per-basepair
per-genome rate that mutations with selection coefficient $s$ enter segments
with feature class $k(g)$, $L_g$ and $r_g$ are the length in basepairs and
recombination rate per basepair of segment $g$, and the recombination distance
between focal site $x$ and segment $g$ is $d(x, g)$. The functional form of $b$
varies depending on whether the classic BGS model is used, or our form of
Santiago and Caballero's (\citeyear{Santiago2016-mu}) equations (see Section
XXX).

@@

\begin{align}
  {N}_{e,\infty} &= N \exp \left( -\frac{V}{2} Q_\infty^2 \right) \\
  {R} &= \frac{4 U s N_{e,\infty}}{\exp(4 N_{e,\infty} s)-1}\\
\end{align}

@@

and $U = 2\mu L$ and $M = r_\text{BP} L$ as above. In this section, we assume
$\mu$ is the mutation rate to variants with heterozygous selection coefficient
$s$ to simplify notation; later, $\mu$ refers to the genome-wide average
mutation rate and the mutation rate is scaled by the DFE. Solving this system
of two equations for each segment also gives the total substitution or
``ratchet" rate per generation per segment ($\widetilde{R}$). The per-basepair
substitution rate of a segment with mutation rate $\mu$ and selection
coefficient $s$ is $\lambda_d(\mu, s) = \nicefrac{\widetilde{R}}{L}$

@@

In practice, to fit such a model, we must discretize both $\mu(s, k(g))$ and
$s$. Our models use a fixed $m_s$-element grid of selection coefficients, and
mutation rates are on a fixed $m_\mu$-element grid, which is later interpolated
over during the maximum likelihood optimization. The number of annotation
features $K$ is variable and set by the annotation data specified. There are
two ways to parameterize the mutation rates and DFE for all annotation classes
in background selection models. First, for the \emph{free-mutation}
parameterization, each annotation feature class has a free mutation rate
parameter for each of these selection coefficients (as in equation
\eqref{suppeqn:b}).

In the free-mutation parameterization, there are $1 + K \times m_s$ for $\pi_0$
and the free mutation rate parameters, where $K$ is the number of annotation
features. The MLE optimization for this approach is unconstrained, though
$\pi_0$ is bounded to be within $10^{-5} \le \pi_0 \le 10^{-1}$ and mutation
rates are bounded within $10^{-11} \le \mu \le 10^{-7}$. The BGS model requires
a mutation rate for each selection coefficient in the grid, for each annotation
class (here, as an example, CDS, UTRs, and phastcons), which are stored in the
$m_s \times K$ matrix $\mathbf{M}_\text{F}$,

@@

which has $2 + K \times (m_s-1)$ free parameters, since each column must sum to
one. This imposes the constraint that $\sum_i w_{i,k} = 1$, requiring
constrained MLE optimization.

@@

After rescaling each segment $N_e$ by the local predicted reduction
$\hat{B}(x)$, the ratchet rates and $B'$ values are re-calculated.

The rescaled $B'$ calculation outputs a $m_\mu \times m_s \times S$
multidimensional array $\mathbf{R}$ of ratchet rates per segment. These are
rescaled by the segment lengths $\mathbf{l}$, giving the per-basepair mutation
rate array. The per-basepair ratchet rate given maximum likelihood estimates of
$\hat{\mu}$ and $\widehat{\mathbf{W}}$ is used to predicted the ratchet rate
for each segment $g$. The DFE estimate for feature $k$ is a column vector of
$\widehat{\mathbf{W}}$, i.e. $\widehat{\mathbf{w}}_k$

\begin{align}
  \lambda_D(g) = 
\end{align}

The predicted ratchet rate for segment $g$ $\lambda_d(g)$ is based on these
values.
